Contrasting snow algae communities in the mountains of British Columbia, Canada

### 2019-12-23 manuscript/manuscript.Rmd
crunched out a rough sketch of intro...

Snow algae are cold-adapted microalgae that bloom on melting summer snow throughout alpine and polar regions. Snow algae darken the snow surface and decrease albedo[@lutz_biogeography_2016;onuma_temporal_2019]---by one estimate, enough to substantially increase snow melt[@ganey_role_2017]. Estimates of global coverage are lacking, but snow algae blooms can cover many square kilometers of glacier summer snow[@ganey_role_2017], are present on snow overlying sea ice[@gradinger_snow_1996], and are reported from every continent[@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019], yet their impact on biogeochemical cycles and food webs has not been explored. Snow algae has been observed throughout history from the ancient Romans[plinius_naturalis_1096] to Charles Darwin[@darwin_voyage_1959], yet despite nearly a century of microscopic study[@leya_snow_2013] we still lack the most basic understanding of snow algae blooms. How do algae colonise the surface of white snow in spring? How do cells propagate in this harsh environment? Are snow algae obligate symbionts of bacteria, like many other microalgae?[@seymour_zooming_2017] How do snow algae aquire nutrients? Central to answering these questions, we first need to define what species of algae cause snow algae blooms.

@@

<!-- 1. are there distinct communities of snow algae? -->
<!-- rationale: lit suggests chlaino, chloro, sanguina dominant, is there much overlap between the communities? Differences between high and low elevation in microscopy, are there genetic differences? This is an important first step before we measure habitat variables, will allow us to choose which variables to measure... -->
<!-- significance. different community, different ecology and or life history. and the snow algae are the primary producers of the community. define the study system.  -->
<!-- hypothesis: there are distinct communities of snow algae-->
<!-- methods: cell count, 18s, and rbcL amplicon sequencing -->
<!-- results: there is much overlap between communities, but many with dominant species. Some chloromonas, raphi only found with sanguina. Chlainomonas widespread. overall, no, there are not sharply delineated communities. -->
<!-- new questions raised: what habitat differences are found with different communities? is bass becking applicable here? what ecological differences, life history differences characterise the different communities?  -->

Our knowledge of snow algae has deepened since the early days of morphology-based classification, yet still snow algae community composition at the species level remains unknown. Morphology can be misleading when identifying microalgae: the same species can look completely different at different life stages; conversely, different species can look nearly identical.

The first efforts to use genetic sequencing to identify snow algae relied on Sanger sequencing of isolates. This approach has been fruitful *Chloromonas*[@matsuzaki_taxonomic_2019]

@@

Snow algae are the primary producer in a little-studied microbiome that includes protists, fungi, bacteria, and viruses[@naff_phylogeny_2013;@lutz_biogeography_2016;@hamilton_primary_2017;@terashima_microbial_2017;@hisakawa_metagenomic_2015].

>>>>

Snow algae are cold-adapted microalgae that bloom on melting summer snow throughout alpine and polar regions. Snow algae darken the snow surface and decrease albedo[@lutz_biogeography_2016;onuma_temporal_2019]---by one estimate, enough to substantially increase snow melt[@ganey_role_2017]. Estimates of global coverage are lacking, but snow algae blooms can cover many square kilometers of glacier summer snow[@ganey_role_2017], are present on snow overlying sea ice[@gradinger_snow_1996], and are reported from every continent[@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019], yet their impact on biogeochemical cycles and food webs has not been explored. Snow algae has been observed throughout history from the ancient Romans[plinius_naturalis_1096] to Charles Darwin[@darwin_voyage_1959], yet despite nearly a century of microscopic study[@leya_snow_2013] we still lack the most basic understanding of snow algae blooms. How do algae colonise the surface of white snow in spring? How do cells propagate in this harsh environment? Are snow algae obligate symbionts of bacteria, like many other microalgae?[@seymour_zooming_2017] How do snow algae aquire nutrients? Central to answering these questions, we first need to define what species and communities of algae that cause snow algae blooms.

What are snow algae? Leya defines snow algae as algae that can proliferate on the snow, not just algae that are present on the snow. Thus, cultured isolates taken from the snow surface are not necessarily snow algae by this definition, they could be not actively growing in the snow, and may not be the same species as the cells that are most abundant in snow samples. Many snow algae species are reported from Chlorophyceae, including the red cells of Sanguina and Chlainomonas, and Chloromonas, but snow algae in other families have been reported including Trebouxiophyceae, and Chrysophyceae.

Lutz shows that all snow algae communities look the same in Arctic red snow, dominated by "uncultured Chlamydomonadaceae" with lesser amounts of Chloromonas and Trebouxiophyceae. The implication that red snow communities are similar is supported by Prochazkova, who through phylogenetic work suggest that the uncultured Chlamydomonadaceae of Lutz, and the Chlamydomonas nivalis of previous work should be regarded as a novel genus: Sanguina. Chlamydomonas nivalis haplotypes are globally distributed, possible leading to the false impression that snow algae communities are similar across the globe.

However there are many other types of snow algae besides Sanguina, including field sequences from red snow that have not been assigned taxonomy [prochazkova].

Although red snow algae are hypothesized to undergo a green life phase, no one has found green patches of snow with sequence matching red snow algae, nor cultured red snow algae in the lab. Field samples of green snow algae contain algae from taxa not known to produce red pigment, such as Trebouxiophyceae [lutz_omics], Chrysophyceae, or Chloromonas[soto, japan]. The latter, Chloromonas, can produce secondary carotenoids[that leya paper, polyptera]. There are  and patches of Bronze snow microscopically dominated by cells described as "Chloromonas nivalis" and "Chloromonas brevispina", found below treeline in microscopy surveys of the American southwest, and in the Giant Mtns of Czech Republic.

There are a diversity of species of Chloromonas, which species are found where? Lutz and Terashima suggested that Chloromonas polyptera is globally distributed in Arctic, and in Japan, but 18S may not be sufficient to resolve closely related species, of which there are many of CHloromonas.

Red snow can also be dominated by different communities in alpine areas, dominated by the large red quadriflagellate cells of Chlainomonas, which half the size of Sanguina, and although they are coloured by the same pigment are closely related to Chloromonas. Chlainomonas has not been detected in 18S amplicon sequencing studies of snow algae to date, whether it is found alongside other snow algae, or whether it is only found in relatively "pure" communities remains unknown.

Several studies have found that Chlorophyceae is the main one, although Trebouxiophyceae, and Chrysophyceae also grow in snow, but no studies have found them growing in red snow. Whether these types of snow algae are as widespread as others

Motility
Secondary carotenoids

is there overlap between the diferent types of snow algae communities?

@@

<!-- are there distinct communities of snow algae? -->
<!-- rationale: lit suggests chlaino, chloro, sanguina dominant, is there much overlap between the communities? Differences between high and low elevation in microscopy, are there genetic differences? This is an important first step before we measure habitat variables, will allow us to choose which variables to measure... -->
<!-- significance. different community, different ecology and or life history. and the snow algae are the primary producers of the community. define the study system.  -->
<!-- hypothesis: there are distinct communities of snow algae-->
<!-- methods: cell count, 18s, and rbcL amplicon sequencing -->
<!-- results: there is much overlap between communities, but many with dominant species. Some chloromonas, raphi only found with sanguina. Chlainomonas widespread. overall, no, there are not sharply delineated communities. -->
<!-- new questions raised: what habitat differences are found with different communities? is bass becking applicable here? what ecological differences, life history differences characterise the different communities?  -->

@@
